---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  comment = "#>",
  fig.path = "man/figures/README-",
  dpi = 300
)
```

# brease: Causally Sound Priors for Binary Experiments

<!-- badges: start -->
[![R-CMD-check](https://github.com/carloscinelli/brease/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/carloscinelli/brease/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `brease` R package implements the **BREASE** (**B**aseline **R**isk, **E**fficacy, and **A**dverse **S**ide **E**ffects) framework for Bayesian analysis of randomized controlled trials with binary treatment and binary outcome, as described in:

> Irons, N.J. and Cinelli, C. (2025). [Causally Sound Priors for Binary Experiments](https://doi.org/10.1214/25-BA1506). *Bayesian Analysis*.

## Installation

You can install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("carloscinelli/brease")
```

## Basic Usage

```{r}
# loads package
library(brease)

# loads data
data("aspirin")

# Aspirin and Myocardial Infarction (Physicians' Health Study)
# 26 MIs in control (N = 11,034), 10 MIs in treatment (N = 11,037)
result <- brease(y0 = aspirin$y0, y1 = aspirin$y1,
                 N0 = aspirin$N0, N1 = aspirin$N1)
result
```

```{r summary}
summary(result)
```

### Posterior Plots

```{r posterior-plots, fig.width = 8, fig.height = 3.5}
plot(result, which = c("theta0", "theta1", "eta_e", "eta_s"))
```

### Sensitivity Analysis

We can examine how the Bayes factor changes across different prior specifications using contour plots:

```{r sensitivity, fig.width = 5, fig.height = 5}
sens <- bf_seq(y0 = aspirin$y0, y1 = aspirin$y1,
               N0 = aspirin$N0, N1 = aspirin$N1,
               mus_seq = seq(0.01, 0.99, length = 15),
               mue_seq = seq(0.01, 0.99, length = 15))
contour_bf(sens)
```

## Overview

The BREASE framework parameterizes the likelihood of a binary experiment in terms of three clinically meaningful causal quantities:

- **Baseline risk** ($\theta_0$): probability of the event without treatment
- **Efficacy** ($\eta_e$): fraction of would-be cases prevented by treatment
- **Side-effect risk** ($\eta_s$): fraction of would-be non-cases harmed by treatment

The package provides:

- **Exact posterior sampling** and analytic marginal likelihoods/Bayes factors
- **Optional MCMC backends** via Stan or JAGS
- **Sensitivity analysis** tools (contour plots, side-effect sensitivity)
- **Comparison methods**: Independent Beta and Logit-Transform priors
- **Bounds** on probabilities of causation

## Documentation

For a detailed introduction, see the [package vignette](https://carloscinelli.com/brease/articles/brease.html).

## Citation

If you use this package, please cite:

```
Irons, N.J. and Cinelli, C. (2025). Causally Sound Priors for Binary
Experiments. Bayesian Analysis. doi:10.1214/25-BA1506
```
