---
title: "An introduction to the BREASE framework"
author: "Nicholas Irons and Carlos Cinelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brease}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

The `brease` package implements the BREASE (**B**aseline **R**isk, **E**fficacy, and **A**dverse **S**ide **E**ffects) framework for Bayesian analysis of randomized controlled trials with binary treatment and binary outcome, as described in [Irons and Cinelli (2025)](https://doi.org/10.1214/25-BA1506).

## The BREASE Parameterization

Consider a two-arm binary experiment where:

- $N_0$ subjects receive control and $y_0$ have the event
- $N_1$ subjects receive treatment and $y_1$ have the event

The BREASE framework parameterizes the model in terms of three causal quantities:

- **Baseline risk** $\theta_0 = P(Y(0) = 1)$: the probability of the event without treatment
- **Efficacy** $\eta_e$: the fraction of would-be cases that are prevented by treatment
- **Side-effect risk** $\eta_s$: the fraction of would-be non-cases that are harmed by treatment

The treatment group risk is then:
$$\theta_1 = \theta_0(1 - \eta_e) + (1 - \theta_0)\eta_s$$

## Specifying Priors

Independent Beta priors are placed on the three causal parameters:

- $\theta_0 \sim \text{Beta}(\mu_0 n_0, (1-\mu_0)n_0)$
- $\eta_e \sim \text{Beta}(\mu_e n_e, (1-\mu_e)n_e)$
- $\eta_s \sim \text{Beta}(\mu_s n_s, (1-\mu_s)n_s)$

The default prior `BREASE(1/2, 0.3, 0.3; 2, 1, 1)` places uniform marginal priors on the outcome probabilities and centers both efficacy and side-effect risk at 0.3 with prior sample size 1.

```{r prior}
library(brease)

# Simulate from the default prior
prior_samps <- sim_brease(10000)
plot(prior_samps$theta0, prior_samps$theta1,
     pch = ".", col = adjustcolor("steelblue", 0.3),
     xlab = expression(theta[0]), ylab = expression(theta[1]),
     main = "BREASE Default Prior")
abline(0, 1, lty = 2, col = "red")
```

## Example: Aspirin and Myocardial Infarction

The Physicians' Health Study (1989) randomized 22,071 male physicians to aspirin or placebo and observed myocardial infarctions (MI).

```{r aspirin}
# Data: 26 MIs in control (N=11,034), 10 MIs in treatment (N=11,037)
result <- brease(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037)
result
```

The posterior summaries show the risk ratio and efficacy fraction, and the Bayes factor quantifies the evidence for a treatment effect.

## Sensitivity Analysis

We can examine how the Bayes factor changes across different prior specifications:

```{r sensitivity, fig.width=5, fig.height=5}
sens <- bf_seq(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037,
               mus_seq = seq(0.01, 0.99, length = 15),
               mue_seq = seq(0.01, 0.99, length = 15))
contour_bf(sens)
```

## Comparison with Other Methods

The package also implements the Independent Beta (IB) and Logit-Transform (LT) approaches for comparison:

```{r comparison}
# Independent Beta
ib_result <- brease_ib(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037)
ib_result
```

## COVID Vaccine Trial

As a second example, consider the Pfizer COVID-19 vaccine trial:

```{r covid}
result_covid <- brease(y0 = 169, y1 = 9, N0 = 20172, N1 = 19965)
result_covid
```

The very strong Bayes factor reflects the overwhelming evidence for vaccine efficacy.

## References

Irons, N.J. and Cinelli, C. (2025). Causally Sound Priors for Binary Experiments. *Bayesian Analysis*. doi:10.1214/25-BA1506
