---
title: "Replication of Irons and Cinelli (2025)"
author: "Nicholas Irons and Carlos Cinelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replication of Irons and Cinelli (2025)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%"
)
set.seed(2025)
```

This vignette replicates the main results and figures from:

> Irons, N.J. and Cinelli, C. (2025). Causally Sound Priors for Binary Experiments. *Bayesian Analysis*. doi:[10.1214/25-BA1506](https://doi.org/10.1214/25-BA1506)

Replication code for the paper is also available at <https://github.com/njirons/causally-sound>.

```{r load}
library(brease)
```

## 1. Prior Specification

The default BREASE prior is $\text{BREASE}(1/2, 0.3, 0.3; 2, 1, 1)$, which places:

- A uniform (Beta(1,1)) prior on baseline risk $\theta_0$
- A Beta(0.3, 0.7) prior on efficacy $\eta_e$ (centered at 0.3)
- A Beta(0.3, 0.7) prior on side-effect risk $\eta_s$ (centered at 0.3)

The induced prior on $(\theta_0, \theta_1)$ exhibits dependence between the treatment arms.

```{r prior-joint, fig.width=7, fig.height=4}
par(mfrow = c(1, 3), mar = c(4, 4, 2.5, 1))

# BREASE prior
brease_samps <- sim_brease(50000)
plot(brease_samps$theta0, brease_samps$theta1,
     pch = ".", col = adjustcolor("steelblue", 0.2),
     xlim = c(0, 1), ylim = c(0, 1),
     xlab = expression(theta[0]), ylab = expression(theta[1]),
     main = "BREASE Prior")
abline(0, 1, lty = 2, col = "red")

# Independent Beta prior
ib_samps <- sim_ib(50000)
plot(ib_samps$theta0, ib_samps$theta1,
     pch = ".", col = adjustcolor("firebrick", 0.2),
     xlim = c(0, 1), ylim = c(0, 1),
     xlab = expression(theta[0]), ylab = expression(theta[1]),
     main = "Independent Beta Prior")
abline(0, 1, lty = 2, col = "red")

# Logit-Transform prior
lt_samps <- sim_lt(50000)
plot(lt_samps$theta0, lt_samps$theta1,
     pch = ".", col = adjustcolor("darkgreen", 0.2),
     xlim = c(0, 1), ylim = c(0, 1),
     xlab = expression(theta[0]), ylab = expression(theta[1]),
     main = "Logit-Transform Prior")
abline(0, 1, lty = 2, col = "red")
```

## 2. Aspirin and Myocardial Infarction

### 2.1 Data

The Physicians' Health Study (1989) randomized 22,071 male physicians to aspirin or placebo and observed fatal myocardial infarctions (MI):

```{r aspirin-data}
data(aspirin)
aspirin
```

### 2.2 BREASE Analysis

```{r aspirin-brease}
# Default prior: BREASE(1/2, 0.3, 0.3; 2, 1, 1)
result <- brease(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037)
result
```

Posterior risk ratio (as percentage):

```{r aspirin-rr}
round(result$post.summaries["rr", ] * 100, 2)
```

Bayes factor in favor of treatment effect:

```{r aspirin-bf}
round(result$bayes.factors$bf10, 2)
```

With a more skeptical prior ($\mu_e = \mu_s = 0.2$), the 95% credible interval for the risk ratio already includes 1:

```{r aspirin-skeptical}
result_02 <- brease(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037,
                     mue = 0.2, mus = 0.2)
round(result_02$post.summaries["rr", ] * 100, 2)
```

### 2.3 Posterior Densities

```{r aspirin-posterior, fig.width=7, fig.height=4}
plot(result, which = c("theta0", "theta1", "eta_e", "eta_s"))
```

```{r aspirin-posterior-effects, fig.width=7, fig.height=4}
plot(result, which = c("rr", "ef", "or", "rd"))
```

### 2.4 Bounds on Probabilities of Causation

```{r aspirin-bounds}
brease_bounds(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037)
```

### 2.5 Sensitivity Analysis

```{r aspirin-side-plot, fig.width=5, fig.height=5}
side_plot(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037)
```

```{r aspirin-contour, fig.width=5, fig.height=5}
sens <- bf_seq(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037,
               mus_seq = seq(0.01, 0.99, length = 20),
               mue_seq = seq(0.01, 0.99, length = 20))
contour_bf(sens)
```

### 2.6 Comparison: Independent Beta

```{r aspirin-ib}
# Default IB with a = b = 1 (uniform)
ib_result <- brease_ib(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037)
round(ib_result$post.summaries["rr", ] * 100, 2)
cat("BF01 (evidence for null):", round(ib_result$bayes.factors$bf01, 2), "\n")

# IB with a = b = 17
ib_result_17 <- brease_ib(y0 = 26, y1 = 10, N0 = 11034, N1 = 11037,
                            a = c(17, 17), b = c(17, 17))
round(ib_result_17$post.summaries["rr", ] * 100, 2)
```

## 3. COVID-19 Vaccine Trial (Pfizer)

### 3.1 Data

The Pfizer-BioNTech COVID-19 vaccine trial (Polack et al., 2020) randomized 40,137 participants:

```{r covid-data}
data(pfizer)
pfizer
```

### 3.2 BREASE Analysis

```{r covid-brease}
result_covid <- brease(y0 = 169, y1 = 9, N0 = 20172, N1 = 19965)
result_covid
```

Posterior efficacy fraction (as percentage):

```{r covid-ef}
round(result_covid$post.summaries["ef", ] * 100, 2)
```

Bayes factor:

```{r covid-bf}
result_covid$bayes.factors$bf10
```

### 3.3 Monotone (No-Harm) Model

The monotone model constrains $\eta_s = 0$ (no side effects):

```{r covid-mono}
result_mono <- brease(y0 = 169, y1 = 9, N0 = 20172, N1 = 19965, mono = TRUE)
round(result_mono$post.summaries["ef", ] * 100, 2)
result_mono$bayes.factors$bf10
```

### 3.4 Posterior Densities

```{r covid-posterior, fig.width=7, fig.height=4}
plot(result_covid, which = c("theta0", "theta1", "eta_e", "eta_s"))
```

```{r covid-posterior-effects, fig.width=7, fig.height=4}
plot(result_covid, which = c("rr", "ef"))
```

### 3.5 Bounds on Probabilities of Causation

```{r covid-bounds}
brease_bounds(y0 = 169, y1 = 9, N0 = 20172, N1 = 19965)
```

### 3.6 Sensitivity Analysis

```{r covid-side-plot, fig.width=5, fig.height=5}
side_plot(y0 = 169, y1 = 9, N0 = 20172, N1 = 19965,
          mus_seq = seq(0.01, 0.99, length = 25),
          log_scale = TRUE)
```

```{r covid-contour, fig.width=5, fig.height=5}
sens_covid <- bf_seq(y0 = 169, y1 = 9, N0 = 20172, N1 = 19965,
                     mus_seq = seq(0.01, 0.99, length = 20),
                     mue_seq = seq(0.01, 0.99, length = 20))
contour_bf(sens_covid)
```

### 3.7 Comparison: Independent Beta

```{r covid-ib}
ib_covid <- brease_ib(y0 = 169, y1 = 9, N0 = 20172, N1 = 19965)
round(ib_covid$post.summaries["ef", ] * 100, 2)
ib_covid$bayes.factors$bf10
```

## 4. NEJM Dataset: Comparing Methods

### 4.1 Data

We use 39 clinical trials from the New England Journal of Medicine to compare BREASE, Independent Beta (IB), and Logit-Transform (LT) methods.

```{r nejm-data}
data(nejm)
head(nejm)
```

### 4.2 Log Marginal Likelihoods

We compute the log marginal likelihood under $H_1$ (treatment effect) for each study using all three methods.

```{r nejm-lml, fig.width=7, fig.height=5}
# BREASE log marginal likelihoods (default prior)
lml_brease_vals <- sapply(1:nrow(nejm), function(i) {
  lml_brease(y0 = nejm$y1[i], y1 = nejm$y2[i],
             N0 = nejm$n1[i], N1 = nejm$n2[i],
             mu0 = 0.5, n0 = 2, mue = 0.3, ne = 1, mus = 0.3, ns = 1)
})

# IB log marginal likelihoods (a = b = 1)
lml_ib_vals <- sapply(1:nrow(nejm), function(i) {
  lml_brease(y0 = nejm$y1[i], y1 = nejm$y2[i],
             N0 = nejm$n1[i], N1 = nejm$n2[i],
             mu0 = 0.5, n0 = 2, mue = 0.5, ne = 2, mus = 0.5, ns = 2)
})
# For IB, use the analytic IB formula via lml_brease_h0 helper
# (We compute IB LML using the independent Beta-Binomial formula)
lml_ib_vals <- sapply(1:nrow(nejm), function(i) {
  a <- 1; b <- 1
  lchoose(nejm$n1[i], nejm$y1[i]) + lchoose(nejm$n2[i], nejm$y2[i]) +
    lbeta(nejm$y1[i] + a, nejm$n1[i] - nejm$y1[i] + b) - lbeta(a, b) +
    lbeta(nejm$y2[i] + a, nejm$n2[i] - nejm$y2[i] + b) - lbeta(a, b)
})

cols <- c("firebrick", "steelblue", "darkgreen")

par(mar = c(5.2, 5.2, 2.1, 2.1))
ylim <- range(c(lml_ib_vals, lml_brease_vals))
plot(1:nrow(nejm), lml_ib_vals,
     col = cols[1], pch = 18, cex = 1.4,
     axes = FALSE,
     xlab = "Study ID", ylab = "",
     ylim = ylim,
     font.main = 1)
axis(1, at = c(1, seq(5, 35, 5), 39))
axis(2, las = 2, at = round(seq(-19.5, -5, len = 7), 1))
mtext(text = "Log Marginal Likelihood", side = 2, line = 3.5)
points(1:nrow(nejm), lml_brease_vals, pch = 17, cex = 1.2, col = cols[3])
legend("topleft",
       col = cols[c(1, 3)],
       pch = c(18, 17),
       legend = c("IB", "BREASE"),
       horiz = TRUE, bty = "n")
```

### 4.3 Bayes Factors

```{r nejm-bf, fig.width=7, fig.height=5}
# Compute BF01 (evidence for H0) for each study
# BREASE
bf_brease_vals <- sapply(1:nrow(nejm), function(i) {
  exp(bf_brease(y0 = nejm$y1[i], y1 = nejm$y2[i],
                N0 = nejm$n1[i], N1 = nejm$n2[i],
                mu0 = 0.5, n0 = 2, mue = 0.3, ne = 1, mus = 0.3, ns = 1))
})

# IB (analytic)
bf_ib_vals <- sapply(1:nrow(nejm), function(i) {
  a <- 1; b <- 1
  # BF01 = p(data | H0) / p(data | H1)
  lml_h0 <- lchoose(nejm$n1[i] + nejm$n2[i], nejm$y1[i] + nejm$y2[i]) +
    lbeta(nejm$y1[i] + nejm$y2[i] + a, nejm$n1[i] + nejm$n2[i] - nejm$y1[i] - nejm$y2[i] + b) -
    lbeta(a, b)
  lml_h1 <- lchoose(nejm$n1[i], nejm$y1[i]) + lchoose(nejm$n2[i], nejm$y2[i]) +
    lbeta(nejm$y1[i] + a, nejm$n1[i] - nejm$y1[i] + b) - lbeta(a, b) +
    lbeta(nejm$y2[i] + a, nejm$n2[i] - nejm$y2[i] + b) - lbeta(a, b)
  exp(lml_h0 - lml_h1)
})

# BREASE with sensitivity (mu from 0.2 to 0.7)
seq_mu <- seq(0.2, 0.7, length.out = 50)
mu_def <- which.min(abs(seq_mu - 0.3))

bf_brease_sens <- sapply(seq_mu, function(a) {
  sapply(1:nrow(nejm), function(i) {
    exp(bf_brease(y0 = nejm$y1[i], y1 = nejm$y2[i],
                  N0 = nejm$n1[i], N1 = nejm$n2[i],
                  mu0 = 0.5, n0 = 2, mue = a, ne = 1, mus = a, ns = 1))
  })
})

# IB sensitivity (a from 1 to 5)
seq_a <- seq(1, 5, length.out = 50)
bf_ib_sens <- sapply(seq_a, function(a) {
  sapply(1:nrow(nejm), function(i) {
    aa <- a; bb <- a
    lml_h0 <- lchoose(nejm$n1[i] + nejm$n2[i], nejm$y1[i] + nejm$y2[i]) +
      lbeta(nejm$y1[i] + nejm$y2[i] + aa, nejm$n1[i] + nejm$n2[i] - nejm$y1[i] - nejm$y2[i] + bb) -
      lbeta(aa, bb)
    lml_h1 <- lchoose(nejm$n1[i], nejm$y1[i]) + lchoose(nejm$n2[i], nejm$y2[i]) +
      lbeta(nejm$y1[i] + aa, nejm$n1[i] - nejm$y1[i] + bb) - lbeta(aa, bb) +
      lbeta(nejm$y2[i] + aa, nejm$n2[i] - nejm$y2[i] + bb) - lbeta(aa, bb)
    exp(lml_h0 - lml_h1)
  })
})

# Plot
cols <- c("firebrick", "darkgreen")

# Create alpha gradients
alphas_ib <- c(1, seq(0.05, 0.01, length.out = 49))
cols_ib <- sapply(alphas_ib, function(alpha) adjustcolor(cols[1], alpha.f = alpha))

alphas_brease <- c(seq(0.01, 0.07, length.out = mu_def - 1),
                   1,
                   seq(0.07, 0.01, length.out = 50 - mu_def))
cols_brease <- sapply(alphas_brease, function(alpha) adjustcolor(cols[2], alpha.f = alpha))

par(mar = c(5.1, 4.1, 2.1, 2.1))
plot(1:39, bf_ib_sens[, 1], pch = 18, axes = FALSE, cex = 1.4, log = "y",
     xlab = "Study ID", ylab = "", ylim = c(1, 200), col = cols[1],
     font.main = 1)
axis(1, at = c(1, seq(5, 35, 5), 39))
axis(2, las = 2, at = c(1, 3, 10, 20, 100, 150, 200))
mtext(text = "Evidence for H0", side = 2, line = 2)

for (i in 2:50) {
  points(bf_ib_sens[, i], pch = 18, cex = 1.4, col = cols_ib[i])
}
for (i in 1:50) {
  points(bf_brease_sens[, i], pch = 17, cex = 1.2, col = cols_brease[i])
}

# Legend annotation
points_x <- seq(33, 37.5, length.out = 50)
text(30, 200, "IB", cex = 1)
points(points_x, rep(200, 50), pch = 18, cex = 1.4, col = cols_ib)
text(31.75, 200, "1", cex = 0.9)
text(38.75, 200, "5", cex = 0.9)

ydir <- exp(log(130) - (log(200) - log(130)))
text(28.5, ydir, "BREASE", cex = 1)
points(points_x, rep(ydir, 50), pch = 17, cex = 1.2, col = cols_brease)
text(31.7, ydir, ".2", cex = 0.9)
text(38.75, ydir, ".7", cex = 0.9)
```

### 4.4 Empirical Bayes BREASE

Following the paper's Supplement A (Section 5.3), we use data-driven bounds on the probabilities of causation to set the BREASE prior:

```{r nejm-eb, fig.width=7, fig.height=5}
# Compute empirical Bayes hyperparameters using bounds midpoints
eb_mue <- sapply(1:nrow(nejm), function(i) {
  bounds <- brease_bounds(y0 = nejm$y1[i], y1 = nejm$y2[i],
                          N0 = nejm$n1[i], N1 = nejm$n2[i])
  bounds$eta_e_mid
})

eb_mus <- sapply(1:nrow(nejm), function(i) {
  bounds <- brease_bounds(y0 = nejm$y1[i], y1 = nejm$y2[i],
                          N0 = nejm$n1[i], N1 = nejm$n2[i])
  bounds$eta_s_mid
})

# BREASE-EB Bayes factors
bf_eb_vals <- sapply(1:nrow(nejm), function(i) {
  exp(bf_brease(y0 = nejm$y1[i], y1 = nejm$y2[i],
                N0 = nejm$n1[i], N1 = nejm$n2[i],
                mu0 = 0.5, n0 = 2,
                mue = eb_mue[i], ne = 1,
                mus = eb_mus[i], ns = 1))
})

# BREASE-EB log marginal likelihoods
lml_eb_vals <- sapply(1:nrow(nejm), function(i) {
  lml_brease(y0 = nejm$y1[i], y1 = nejm$y2[i],
             N0 = nejm$n1[i], N1 = nejm$n2[i],
             mu0 = 0.5, n0 = 2,
             mue = eb_mue[i], ne = 1,
             mus = eb_mus[i], ns = 1)
})

cols <- c("darkgreen", "orange")

par(mfrow = c(1, 2), mar = c(5.2, 5.2, 1, 0.5))

# Left panel: Bayes Factors
plot(1:nrow(nejm), bf_brease_vals,
     col = "darkgreen", pch = 17, cex = 1.2,
     axes = FALSE, log = "y",
     xlab = "Study ID", ylab = "",
     ylim = c(1, 20),
     font.main = 1)
axis(1, at = c(1, seq(5, 35, 5), 39))
axis(2, las = 2, at = c(1, 3, 10, 20))
mtext(text = "Evidence for H0", side = 2, line = 2.5)
points(1:nrow(nejm), bf_eb_vals, pch = 17, cex = 1.2,
       col = adjustcolor("orange", 0.7))

# Right panel: Log Marginal Likelihoods
ylim_lml <- range(c(lml_brease_vals, lml_eb_vals))
plot(1:nrow(nejm), lml_brease_vals,
     col = "darkgreen", pch = 17, cex = 1.2,
     axes = FALSE,
     xlab = "Study ID", ylab = "",
     ylim = c(-19.5, -4),
     font.main = 1)
axis(1, at = c(1, seq(5, 35, 5), 39))
axis(2, las = 2, at = round(seq(-19.5, -5, len = 7), 1))
mtext(text = "Log Marginal Likelihood", side = 2, line = 3.5)
points(1:nrow(nejm), lml_eb_vals, pch = 17, cex = 1.2,
       col = adjustcolor("orange", 0.7))
legend("bottomright",
       col = c("darkgreen", adjustcolor("orange", 0.7)),
       pch = c(17, 17),
       legend = c("BREASE", "BREASE-EB"),
       bty = "n")
```

## 5. Aspirin Meta-Analysis

### 5.1 Data

```{r meta-data}
data(aspirin_meta)
aspirin_meta
```

### 5.2 No-Pooling Analysis

Each study is analyzed independently:

```{r meta-nopool}
n_strata <- nrow(aspirin_meta)
probs <- c(0.5, 0.025, 0.975)

nopool <- list()
for (j in 1:n_strata) {
  nopool[[j]] <- gibbs_brease(
    n = 10000, y0 = aspirin_meta$y0[j], y1 = aspirin_meta$y1[j],
    N0 = aspirin_meta$N0[j], N1 = aspirin_meta$N1[j],
    mu0 = 0.5, n0 = 2, mue = 0.3, ne = 1, mus = 0.3, ns = 1,
    burn_in = 1000
  )
}

rr_nopool <- do.call("rbind", lapply(nopool, function(x) {
  quantile(x$theta1 / x$theta0, probs)
}))
rr_nopool <- as.data.frame(round(rr_nopool * 100, 2))
rr_nopool <- cbind(Study = aspirin_meta$Study,
                   N = aspirin_meta$N1 + aspirin_meta$N0,
                   rr_nopool)
rr_nopool
```

### 5.3 Complete-Pooling Analysis

Aggregate all evidence as if it were a single study:

```{r meta-complete}
total_y0 <- sum(aspirin_meta$y0)
total_N0 <- sum(aspirin_meta$N0)
total_y1 <- sum(aspirin_meta$y1)
total_N1 <- sum(aspirin_meta$N1)

cat("Total data: y0 =", total_y0, ", N0 =", total_N0,
    ", y1 =", total_y1, ", N1 =", total_N1, "\n")

# Analysis with default mu = 0.3
total_result <- brease(y0 = total_y0, y1 = total_y1,
                       N0 = total_N0, N1 = total_N1,
                       sampler = "gibbs")
round(total_result$post.summaries["rr", ] * 100, 2)

# Bayes factor
cat("BF10:", total_result$bayes.factors$bf10, "\n")
```

With a more skeptical prior ($\mu_e = \mu_s = 0.1$), the 95% CI approximately includes 1:

```{r meta-complete-skeptical}
total_01 <- brease(y0 = total_y0, y1 = total_y1,
                   N0 = total_N0, N1 = total_N1,
                   mue = 0.1, mus = 0.1,
                   sampler = "gibbs", n_samples = 100000)
round(total_01$post.summaries["rr", ] * 100, 2)
```

## 6. Pathological Sampling

This example demonstrates that standard MCMC (JAGS, Stan) can fail with very low side-effect priors, while the exact sampler and data-augmented Gibbs sampler succeed.

```{r pathological, fig.width=7, fig.height=4}
# Data and prior (very low side-effect prior)
y0_p <- 20; N0_p <- 1000; y1_p <- 40; N1_p <- 1000
mu0_p <- 0.5; n0_p <- 2
mue_p <- 0.5; ne_p <- 2   # Uniform on efficacy
mus_p <- 0.01; ns_p <- 1   # Very low side-effect prior

# Exact sampler
exact_fit <- sample_brease(100000, y0 = y0_p, y1 = y1_p,
                           N0 = N0_p, N1 = N1_p,
                           mu0 = mu0_p, n0 = n0_p,
                           mue = mue_p, ne = ne_p,
                           mus = mus_p, ns = ns_p)

# DA Gibbs sampler
gibbs_fit <- gibbs_brease(100000, y0 = y0_p, y1 = y1_p,
                          N0 = N0_p, N1 = N1_p,
                          mu0 = mu0_p, n0 = n0_p,
                          mue = mue_p, ne = ne_p,
                          mus = mus_p, ns = ns_p,
                          burn_in = 1000)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 2))

hist(exact_fit$theta0, freq = FALSE, breaks = 100, ylim = c(0, 150),
     xlab = expression(theta[0]),
     col = "gray", main = NULL, ylab = "Density")
hist(gibbs_fit$theta0, col = adjustcolor("green", 0.3),
     add = TRUE, freq = FALSE, breaks = 100)
legend("topright",
       legend = c("Exact", "DA Gibbs"),
       fill = c("gray", adjustcolor("green", 0.3)),
       bty = "n")

hist(exact_fit$theta1, freq = FALSE, breaks = 100, ylim = c(0, 150),
     xlab = expression(theta[1]),
     col = "gray", main = NULL, ylab = "Density")
hist(gibbs_fit$theta1, col = adjustcolor("green", 0.3),
     add = TRUE, freq = FALSE, breaks = 100)
```

Both the exact and DA Gibbs samplers produce identical posteriors, even in this pathological setting where standard MCMC methods can fail.

## References

- Irons, N.J. and Cinelli, C. (2025). Causally Sound Priors for Binary Experiments. *Bayesian Analysis*. doi:[10.1214/25-BA1506](https://doi.org/10.1214/25-BA1506)
- Steering Committee of the Physicians' Health Study Research Group (1989). Final report on the aspirin component of the ongoing Physicians' Health Study. *New England Journal of Medicine*, 321(3), 129--135.
- Polack, F.P., Thomas, S.J., Kitchin, N. et al. (2020). Safety and Efficacy of the BNT162b2 mRNA Covid-19 Vaccine. *New England Journal of Medicine*, 383(27), 2603--2615.
